Multi-stage programming with functors and monads: eliminating abstraction overhead from generic code

### 2008-02-01 SCP2008/paper/metamonads.tex
take care of a few more comments

\noindent  The types above are
generally lifted twice: once from the value domain |v| to the code
domain |('a,'v) code|, and once more from values to monadic computations
|('p,'v) monad|.

@@

The signature |CONTAINER2D| below specifies that a container must provide
functions |dim1| and |dim2| to extract the dimensions, functions |getL|
to generate container getters, the cloning
generator |copy| and functions that generate code for row and column
swapping. The inclusion of these functions in the signature of all
containers makes it simpler to optimize the relevant functions
depending on the actual representation of the container while not
burdening the users of containers with efficiency details.

@@

\begin{code}
module GenericArrayContainer(Dom:DOMAINL) =
  struct
  module Dom = Dom
  type contr = Dom.v array array (* Array of rows *)
  type 'a vc = ('a,contr) code
  type 'a vo = ('a,Dom.v) code
  let getL x n m = .< (.~x).(.~n).(.~m) >.
  let dim2 x = .< Array.length .~x >.       (* number of rows *)
  let dim1 x = .< Array.length (.~x).(0) >. (* number of cols *)
  \dots
\end{code}
%
The accompanying code includes containers whose
elements are stored in a 1D array, in a
row-wise (C-like) or column-wise (Fortran-like) modes.

@@

Various aspects of our LU generator such as determinant, rank, 
and permutation tracking may need to keep a state during
code generation. For example, the initializing section of the LU
generator will invoke the function |decl| of the determinant aspect so that the
latter may generate let-binding for reference cells that track the
magnitude and the sign of the determinant. The generator of the
row-swapping code will notify the determinant aspect so that the
latter may insert code that flips the sign of the tracked
determinant. The final section of the LU generator will ask the
determinant aspect to generate code producing the final, signed value
of the determinant. Clearly the determinant aspect needs to maintain
state from one invocation of its functions to another; this state should
include the ``names'' of mutable variables accumulating the sign and the
magnitude of the determinant.

@@

The MLton team suggests a better approach: property 
lists~\cite{mlton-proplist}.  A property list, too, represents an
extensible object -- by its dual, a list of polymorphic variants.  The
initial object is just the empty list. MLton's approach does not
literally apply to us however. MetaOCaml's environment classifiers
require type universes parameterized by classifiers. The MLton
approach uses generativity of exceptions or reference cells to
generate property names, i.e., fields of an extensible record. That
would make our aspects stateful modules, which we try to
avoid. Fortunately, OCaml offers a way to build open unions with
polymorphic variants. Each variant is identified by a manifest tag,
e.g., |`Tdet|, and may include a value. The tags are not generative and
provide a form of manifest naming, similar to symbols of
Lisp and Scheme.

>>>>

\noindent  The types above are
generally lifted in two ways: once from the value domain |v| to the code
domain |('a,'v) code|, and once more from values to monadic computations
|('p,'v) monad|.

@@

For our purposes, a container is an abstraction of an $n$-dimensional
vector space, which we here specialize for $n=1,2$.  The $2$ dimensional
case is our main interest, and its signature contains many functions particular
to $n=2$.  For examples, we have rows and columns and operations specialized
for them.  A container explicitly abstracts the underlying representation 
of the data-structure, while offering an interface which is better-suited
to linear algebra.

The signature |CONTAINER2D| below specifies that a container must provide
functions |dim1| and |dim2| to extract the dimensions, functions |getL|
to generate container getters, the cloning
generator |copy| and functions that generate code for row and column
swapping. The inclusion of these functions in the signature of all
containers makes it simpler to optimize the relevant functions
depending on the actual representation of the container while not
burdening the users of containers with efficiency details.  Note that
this module type is also abstract over the actual representation 
|rep| of the generator data-structure.

@@

\begin{code}
module GenericArrayContainer(Dom:DOMAINL) =
  struct
  module Dom = Dom
  type contr = Dom.v array array (* Array of rows *)
  type 'a vc = ('a,contr) code
  type 'a vo = ('a,Dom.v) code
  let getL x n m = .< (.~x).(.~n).(.~m) >.
  let dim2 x = .< Array.length .~x >.       (* number of rows *)
  let dim1 x = .< Array.length (.~x).(0) >. (* number of cols *)
  \dots
\end{code}
%
The accompanying code\cite{metamonadsURL} includes containers whose
elements are stored in a 1D array, in a
row-wise (C-like) or column-wise (Fortran-like) modes.

@@

Various aspects of our LU generator such as determinant, rank, 
and permutation tracking may need to keep a state during
code generation. For example, the initializing section of the LU
generator will invoke a function (|decl|) of the determinant aspect 
which ``declares'' whether tracking the determinant is needed, 
so that the
latter may generate let-binding for reference cells that track the
magnitude and the sign of the determinant. The generator of the
row-swapping code will notify the determinant aspect so that the
latter may insert code that flips the sign of the tracked
determinant. The final section of the LU generator will ask the
determinant aspect to generate code producing the final, signed value
of the determinant. Clearly the determinant aspect needs to maintain
state from one invocation of its functions to another; this state should
include the ``names'' of mutable variables accumulating the sign and the
magnitude of the determinant.

@@

The MLton team suggests a better approach: property 
lists~\cite{mlton-proplist}.  A property list also represents an
extensible object but by its dual, namely a list of polymorphic variants.  The
initial object is just the empty list.  Unfortunately, we cannot apply
MLton's approach litterally to our case.  MetaOCaml's environment classifiers
require type universes parameterized by classifiers. The MLton
approach uses generativity of exceptions or reference cells to
generate property names, i.e., fields of an extensible record. Using this
technique would make our aspects stateful modules, something we do not
want, as we wish our generator to be purely functional [even though it is
generating imperative code].
Fortunately, OCaml offers a way to build open unions with
polymorphic variants. Each variant is identified by a manifest tag,
e.g., |`Tdet|, and may include a value. The tags are not generative and
provide a form of manifest naming, similar to symbols of
Lisp and Scheme.
